Choose a License

### 2024-11-28 choose_license.qmd
Use APA citation style
Mention regulation by funder

---
title: "Choose a License"
lang: en
bibliography: literature.bib
format:
  html:
    mermaid:
      theme: neutral
engine: knitr
---

@@

First, if you adapt (i.e., modify, build on) a work by others you need to determine if it is provided to you under a free/open license.[You can determine whether a license is free/open by searching for its name in the [SPDX License List](https://spdx.org/licenses/) and looking for at least one `Y` in the two columns _FSF Free/Libre?_ and _OSI Approved?_]{.aside} If yes, we recommend you to make your contribution available under the same license.^[Copyleft licenses even require you to choose the same or a compatible license.] For example, if you adapt code published in another paper, choose the same license for your modifications. The same applies if there are strong community norms to use a particular free/open license.^[Of course, this is only a heuristic and there might be good reasons to deviate from community norms.] Importantly, as discussed before, you are generally not allowed to adapt a work _not_ published under a free/open license.

@@

::: {#fig-flowchart-simple}
```{mermaid}
flowchart TB
  start("We want<br>to choose<br>a license.") --"We adapted a work by<br>others shared under a<br>free/open license."--> use_existing_license["<em>Use its license</em>"]
  start --"We created the work<br>entirely by ourselves."--> norm("Community norm<br>regarding license?")

  norm --"Exists"--> follow_existing_norms["<em>Follow that norm</em>"]
  norm --"Does not<br>exist"--> type("Work type?")

@@

- __fonts:__ Copyleft licenses applied to fonts can be a special case: If a font is put under the license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/), any documents containing texts using that font will probably be derivative works and have to be put under the same license if shared. If the intent is that only derived fonts, if published, have to be put under the same license, the [SIL Open Font License\ 1.1](https://openfontlicense.org/) is an appropriate choice. Note, however, that it doesn't require attribution for usage. If no copyleft mechanism is intended, [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) also works for fonts.
- __templates and LaTeX packages:__ If a template or, for this purpose, a LaTeX package, is licensed under a copyleft software license such as the [AGPLv3](https://choosealicense.com/licenses/gpl-3.0/), every work that is a derivative of the template or that uses the LaTeX package has to put under the same license, if shared [@Koppor2016]. And if a document contained source code covered by the GPLv3, the same license would also apply to the document [@MadHatter2022].
- __database content:__ If the work produced from a database (the "output") is a derivative of the content in the database, the output is subject to the restrictions laid out in the license. For example, if geospatial data were to be licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/), all maps produced from the data would likely need to fulfill this license's obligation for unrestricted access (its Anti-DRM provision) if shared [see @Poole2017]. Similarly, following an example from @Matt2009, if one were to choose the copyleft license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) for this purpose, any map that is a derivative of the data would also need to licensed under CC\ BY-SA\ 4.0 (or a compatible license) if shared. If the intention is to only have [derivative databases](https://osmfoundation.org/wiki/Licence/Community_Guidelines/Produced_Work_-_Guideline) under the same license, one might want to choose the [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/) for the database, as it was specifically designed not to apply to works produced from the data in the database. Otherwise [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) is an excellent choice for data.
- __works in the public domain__: If a work is already in the public domain, it should be marked using the [PDM 1.0](https://creativecommons.org/publicdomain/mark/1.0/), rather than applying a waiver such as the [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) (or even another license).
:::

@@

::: {#fig-flowchart-non-data}
```{mermaid}
flowchart TB
  start("We want to choose a<br>license for software,<br>writing, image, audio,<br>or video.") --"We adapted a work by<br>others shared under a<br>free/open license."--> use_existing_license["<em>Use its license</em>"]
  start --"We created the work<br>entirely by ourselves."--> norm("Community norm<br>regarding license?")

  norm --"Exists"--> follow_existing_norms["<em>Follow that norm</em>"]
  norm --"Does not<br>exist"--> type("Work type?")

@@

::: {#fig-flowchart-data}
```{mermaid}
flowchart TB
  start("We want to choose a<br>license for data.") --"We adapted a database by<br>others shared under a<br>free/open license."--> use_existing_license_db["<em>Use its license(s)<br>for content and database</em>"]
  start --"We created a database<br>entirely by ourselves."--> norm("Community norm<br>regarding license?")

  norm --"Exists"--> follow_existing_norms["<em>Follow that norm</em>"]
  norm --"Does not<br>exist"--> existing_license_content("Adapting content<br> by others?")

  subgraph "License for individual data entries (content)"
  existing_license_content --"No, we created the content<br>entirely by ourselves."--> metadata("Entries are facts<br>(like measurements<br>or metadata)?")
  existing_license_content --"Yes, it was<br>shared under a<br>free/open license."--> use_existing_license_content["<em>Use that license</em>"]
  metadata --"Yes"--> cc0_content_metadata["CC0 1.0"]
  metadata --"No"--> choose_license["<em>Consult flowchart for<br>software, writing,<br>image, audio, and video</em>"]
  end

>>>>

---
title: "Choose a License"
lang: en
format:
  html:
    mermaid:
      theme: neutral
engine: knitr
---

@@

First, if you adapt (i.e., modify, build on) a work by others you need to determine if it is provided to you under a free/open license.[You can determine whether a license is free/open by searching for its name in the [SPDX License List](https://spdx.org/licenses/) and looking for at least one `Y` in the two columns _FSF Free/Libre?_ and _OSI Approved?_]{.aside} If yes, we recommend you to make your contribution available under the same license.^[Copyleft licenses even require you to choose the same or a compatible license.] For example, if you adapt code published in another paper, choose the same license for your modifications. The same applies if there are strong community norms to use a particular free/open license^[Of course, this is only a heuristic and there might be good reasons to deviate from community norms.] or if a particular license choice is mandated by your funder. Importantly, as discussed before, you are generally not allowed to adapt a work _not_ published under a free/open license.

@@

::: {#fig-flowchart-simple}
```{mermaid}
flowchart TB
  start("We want<br>to choose<br>a license.") --"We adapted a work by<br>others shared under a<br>free/open license."--> use_existing_license["<em>Use its license</em>"]
  start --"We created the work<br>entirely by ourselves."--> regulation("Other regulation<br>(by community or funder)<br>regarding license?")

  regulation --"Exists"--> follow_existing_norms["<em>Follow that</em>"]
  regulation --"Does not<br>exist"--> type("Work type?")

@@

- __fonts:__ Copyleft licenses applied to fonts can be a special case: If a font is put under the license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/), any documents containing texts using that font will probably be derivative works and have to be put under the same license if shared. If the intent is that only derived fonts, if published, have to be put under the same license, the [SIL Open Font License\ 1.1](https://openfontlicense.org/) is an appropriate choice. Note, however, that it doesn't require attribution for usage. If no copyleft mechanism is intended, [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) also works for fonts.
- __templates and LaTeX packages:__ If a template or, for this purpose, a LaTeX package, is licensed under a copyleft software license such as the [AGPLv3](https://choosealicense.com/licenses/agpl-3.0/), every work that is a derivative of the template or that uses the LaTeX package has to put under the same license, if shared [@Koppor2016]. And if a document contained source code covered by the GPLv3, the same license would also apply to the document [@MadHatter2022].
- __database content:__ If the work produced from a database (the "output") is a derivative of the content in the database, the output is subject to the restrictions laid out in the license. For example, if geospatial data were to be licensed under [CC\ BY\ 4.0](https://creativecommons.org/licenses/by/4.0/), all maps produced from the data would likely need to fulfill this license's obligation for unrestricted access (its Anti-DRM provision) if shared [see @Poole2017]. Similarly, following an example from @Matt2009, if one were to choose the copyleft license [CC\ BY-SA\ 4.0](https://creativecommons.org/licenses/by-sa/4.0/) for this purpose, any map that is a derivative of the data would also need to licensed under CC\ BY-SA\ 4.0 (or a compatible license) if shared. If the intention is to only have [derivative databases](https://osmfoundation.org/wiki/Licence/Community_Guidelines/Produced_Work_-_Guideline) under the same license, one might want to choose the [ODbL\ 1.0](https://opendatacommons.org/licenses/odbl/summary/) for the database, as it was specifically designed not to apply to works produced from the data in the database. Otherwise [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) is an excellent choice for data.
- __works in the public domain__: If a work is already in the public domain, it should be marked using the [PDM 1.0](https://creativecommons.org/publicdomain/mark/1.0/), rather than applying a waiver such as the [CC0\ 1.0](https://creativecommons.org/publicdomain/zero/1.0/) (or even another license).
:::

@@

::: {#fig-flowchart-non-data}
```{mermaid}
flowchart TB
  start("We want to choose a<br>license for software,<br>writing, image, audio,<br>or video.") --"We adapted a work by<br>others shared under a<br>free/open license."--> use_existing_license["<em>Use its license</em>"]
  start --"We created the work<br>entirely by ourselves."--> regulation("Other regulation<br>(by community or funder)<br>regarding license?")

  regulation --"Exists"--> follow_existing_norms["<em>Follow that</em>"]
  regulation --"Does not<br>exist"--> type("Work type?")

@@

::: {#fig-flowchart-data}
```{mermaid}
flowchart TB
  start("We want to choose a<br>license for data.") --"We adapted a database by<br>others shared under a<br>free/open license."--> use_existing_license_db["<em>Use its license(s)<br>for content and database</em>"]
  start --"We created a database<br>entirely by ourselves."--> regulation("Other regulation<br>(by community or funder)<br>regarding license?")

  regulation --"Exists"--> follow_existing_norms["<em>Follow that</em>"]
  regulation --"Does not<br>exist"--> existing_license_content("Adapting content<br> by others?")

  subgraph "License for individual data entries (content)"
  existing_license_content --"No, we created the content<br>entirely by ourselves."--> facts("Entries are facts<br>(like measurements<br>or metadata)?")
  existing_license_content --"Yes, it was<br>shared under a<br>free/open license."--> use_existing_license_content["<em>Use that license</em>"]
  facts --"Yes"--> cc0_content_metadata["CC0 1.0"]
  facts --"No"--> choose_license["<em>Consult flowchart for<br>software, writing,<br>image, audio, and video</em>"]
  end
